The illusory promise of the Aligned Rank Transform

### 2025-06-13 text/10-intro.qmd
updates in Section 3... still ongoing

## Introduction {#intro}
In Human-Computer Interaction (HCI) and various fields within the behavioral sciences, researchers often gather data through user studies. Such data are typically messy, have small sample sizes, and may violate common statistical assumptions, such as the normality assumption. To address these challenges, researchers commonly employ nonparametric tests, which require fewer assumptions about the data. However, while nonparametric tests for simple one-factorial designs are well-established, researchers face challenges in selecting appropriate methods when dealing with multifactorial designs that require testing for both main effects and interactions. The Aligned Rank Transform or ART [@higgins:1990; @Salter:1993; @wobbrock:2011] addresses this problem by bridging the gap between nonparametric tests and ANOVAs. Its popularity in HCI research has surged, facilitated in part by the ARTool toolkit [@wobbrock:2011; @artool], which simplifies the use of the method.

@@

Our goal is to clarify the severity of these issues and understand the potential risks of the method. We present the outcomes of a series of Monte Carlo experiments, following a distinctive simulation methodology grounded in latent variable modeling. This approach enables us to simulate effects consistently across a broad spectrum of distributions, both discrete and continuous, and allows us to address scaling issues in interpreting interactions. To ensure the clarity of our findings and facilitate their reproducibility, we divide the experimental process into multiple smaller experiments, where each experiment focuses on a distinct variable (e.g., sample size, experimental design, variance ratio) or measure (Type I error rate, power, precision of effect size estimates).

@@

ART's *p*-values for the two main effects are surprisingly low, while no other methods shows any substantial evidence for such effects. We also presents the methods' effect size estimates:

@@

The paper is structured as follows. [Section 2](#background) introduces nonparametric tests and rank transformations, and explains how ART is constructed. It also provides a summary of previous experimental results regarding the robustness of ART and other related rank-based procedures, along with a survey of recent studies using ART. [Section 3](#interpretation) investigates issues regarding effect interpretation and introduce our distribution simulation approach. [Section 4](#methodology) outlines our experimental methodology, while [Section 5](#findings) presents our findings. [Section 6](#casestudies) revisits results of previous studies employing ART and  illustrates how its application can lead to erroneous conclusions. [Section 7](#recommendations) offers recommendations for researchers. Finally, [Section 8](#conclusions) concludes the paper.

>>>>

## Introduction {#intro}
In Human-Computer Interaction (HCI) and various fields within the behavioral sciences, researchers often gather data through user studies. Such data are typically messy, have small sample sizes, and may violate common statistical assumptions, such as the normality assumption. To address these challenges, researchers commonly employ nonparametric tests, which require fewer assumptions about the data. However, while nonparametric tests for simple one-factorial designs are well-established, researchers face challenges in selecting appropriate methods when dealing with multifactorial designs that require testing for both main effects and interactions. The Aligned Rank Transform or ART [@higgins:1990; @Salter:1993; @wobbrock:2011] addresses this problem by bridging the gap between nonparametric tests and ANOVAs. Its popularity in HCI research has surged, in part due to the ARTool toolkit [@wobbrock:2011; @artool], which makes the method easy to use. @Oppenlaender:2025 identify the paper by @wobbrock:2011 as major milestone in HCI research, ranking third in terms of citations in the ACM CHI Proceedings (1981-2024), behind Braun and Clark's [-@Braun:2006] article on thematic analysis and Hart and Staveland's [-@hart1988development] article on the NASA-TXL index.

However, given the significant influence of such milestone methods on our community's research outcomes, they warrant a closer scrutiny. Our results suggest that ART's popularity step largely from an ignorance of the method's assumptions. Researchers often treat ART as a general-purpose nonparametric alternative when ANOVA's assumptions are violated, overlooking the fact that ART's alignment mechanism introduces stricter assumptions than ANOVA itself, including a linear relationship between the response variable and the effects, equal variances, and continuous responses.

@@

This paper aims to clarify the severity of these issues and understand the potential risks of the method. We present the outcomes of a series of Monte Carlo experiments, following a distinctive simulation methodology grounded in latent variable modeling. This approach enables us to simulate effects consistently across a broad spectrum of distributions, both discrete and continuous, and allows us to address issues in interpreting main and interaction effects. To ensure the clarity of our findings and facilitate their reproducibility, we divide the experimental process into multiple smaller experiments, where each experiment focuses on a distinct variable (e.g., sample size, experimental design, variance ratio) or measure (Type I error rate, power, precision of effect size estimates).

@@

ART's *p*-values for the two main effects are surprisingly low, while no other method shows any substantial evidence for such effects. We also present the methods' effect size estimates:

@@

The paper is structured as follows. [Section 2](#background) introduces nonparametric tests and rank transformations, and explains how ART is constructed. It also provides a summary of previous experimental results regarding the robustness of ART and other related rank-based procedures, along with a survey of recent studies using ART. [Section 3](#assumptions) clarifies the method's distributional assumptions. [Section 4](#interpretation) investigates issues regarding effect interpretation and introduce our distribution simulation approach. [Section 5](#methodology) outlines our experimental methodology, while [Section 6](#findings) presents our findings. [Section 7](#casestudies) revisits results of previous studies employing ART and  illustrates how its application can lead to erroneous conclusions. [Section 8](#recommendations) offers recommendations for researchers. Finally, [Section 9](#conclusions) concludes the paper.

### 2025-06-13 text/20-background.qmd
updates in Section 3... still ongoing

Our goal is to verify Lüpsen's findings by employing an alternative set of experimental configurations and a distinctly different methodology. In particular, we adopt a latent variable modeling approach that establishes a unified framework for data generation across all distributions. This approach allows us to demonstrate the shortcomings of rank-based approaches in effectively addressing interactions through the lens of removable interactions [@Loftus:1978]. Moreover, it facilitates the assessment of the methods through more suitable generation procedures for ordinal data, as suggested by @Liddell:2018.

With a focus on clarity, we chose to exclusively examine the three rank-based transformations presented earlier: the pure rank transformation (RNK), ART, and INT. While we do not elaborate on the performance of ATS in the main paper, additional experimental results are available in the [appendix](appendix.html). These findings indicate that its performance is comparable to the rank transformation but seems to be inferior to the simpler and more versatile INT. The appendix features additional results regarding the performance of the generalized van der Waerden test, as well as the generalized Kruskal-Wallis and Friedman tests. Our findings show that, at least in balanced designs, these methods suffer from low power issues without demonstrating any clear benefits compared to INT.

>>>>

Our goal is to verify Lüpsen's findings by employing an alternative set of experimental configurations and a distinctly different methodology. In particular, we adopt a latent variable modeling approach that establishes a unified framework for data generation across all distributions. This approach allows us to identify interpretation issues regarding the definition of the null hypothesis, especially for interaction effects, and better explain when and why each method fails. Moreover, it facilitates the assessment of the methods through more suitable generation procedures for ordinal data, as suggested by @Liddell:2018.
Our simulation strategy is similar to that of @elkin:2021 but employs a simpler method for generating data from a latent parameter space. In addition, we evaluate critical scenarios that @elkin:2021 did not consider, particularly the influence of no-null effects and discrete data.

With a focus on clarity, we chose to exclusively examine the three rank-based transformations presented earlier: the pure rank transformation (RNK), ART, and INT. While we do not elaborate on the performance of ATS in the main paper, additional experimental results are available in the [appendix](appendix.html). These findings indicate that its performance is comparable to the rank transformation but seems to be inferior to the simpler and more versatile INT. The appendix features additional results regarding the performance of the generalized van der Waerden test, as well as the generalized Kruskal-Wallis and Friedman tests. Our findings show that, at least in balanced designs, these methods suffer from low power issues in certain situations without demonstrating any clear benefits compared to INT.

### 2025-06-13 text/30-assumptions.qmd
updates in Section 3... still ongoing

### Violations of the linearity assumption
Suppose we studied image-editing tasks (as in our illustrative example), measuring the time participants needed to complete them for two difficulty levels: *easy* and *hard*. If we assume that observed times follow log-normal distributions, the model of @higgins:1990 will generate distributions as the ones in @fig-higgins-model. We observe that the distribution for the hard task is simply shifted to the right, while their shape conveniently remains identical. @Mansouri:1995 have shown that at least for interaction effects, ART remains robust under such distributions. However, how realistic are they?

@@

In these distributions, the mean for had tasks also increases, but this increase is not reflected as a simple global shift in the distribution. Instead, the overall shape of the distribution changes. The model for these distributions is structured as follows:

@@

where $\epsilon_{ijk}$ is normally distributed, and $\theta$ represents a threshold below which response values cannot occur. Unfortunately, ART's alignment mechanism is not designed to handle such distributions. The alignment calculations shown in @fig-ART-explained include various mean terms that are sensitive to distribution differences across the levels of factors, and --- as our results show --- they confounds effects.

We acknowledge that @elkin:2021 diverged from the modeling approach described by @higgins:1990 and @Mansouri:1995 and used generalized linear models as in @eq-generalized-model to evaluate ART. However, their experiments failed to observe that ART confounds effect, simply because they assessed Type I error rates only under scenarios where main effects for all factors were null.

### Heteroscedasticity issues
@myers2012 explain that *"if the underlying distribution of the response variable is not normal, but is a continuous, skewed distribution, such as the lognormal, gamma, or Weibull distribution, we often find that the constant variance assumption is violated,* and in such cases, *"the variance is a function of the mean"* (Pages 54-55). This pattern frequently occurs in studies measuring task-completion times, whether the task is visual, motor, or cognitive. As tasks become more difficult and prolonged, variance tends to increase. Similarly, slower participants generally exhibit greater variance across tasks compared to faster, more practiced users. 
This suggests that ART's assumption of constant variance in skewed distributions is unrealistic.

@@

However, a question remains: Is ART’s failure in handling such models due solely to heteroscedasticity issues? For example, consider the two time distributions in @fig-homoscedastic-model, which --- despite having different shapes --- share the same standard deviation. Would ART perform correctly in this case? While we do not directly address this question in the main paper, we include additional results in the [appendix](appendix.html) showing that as long as distribution shapes differ across levels, ART tends to confound effects. Heteroscedasticity may exacerbate the problem, but it is not its sole cause.

@@

@fig-art-binary illustrates this problem using a simple dataset with a binary variable. In our example, all but one value are zero. Notice how dramatically ART alters the ranks and exaggerates the difference between the two levels of $B$ ($b_1$ vs. $b_2$). This issue persists even when a large number of participants are added and there is only a single value of 1 among thousands of 0s, causing the method to produce arbitrarily low *p*-values. Then, a single value variation (e.g., turning this 1 to 0) will cause ART’s results to drastically change.

>>>>

### Violations of the linearity assumption
Suppose we studied image-editing tasks (as in our illustrative example), measuring the time participants needed to complete them for two difficulty levels: *easy* and *hard*. If we assume that observed times follow log-normal distributions, the model of @higgins:1990 will generate distributions as the ones in @fig-higgins-model. We observe that the distribution for the hard task is simply shifted to the right, while their shape conveniently remains identical. @Mansouri:1995 have shown that at least for interaction effects, ART remains robust under such distributions. However, how realistic are these distributions?

@@

In these distributions, the mean for had tasks also increases, but this increase is not reflected as a simple global shift in the distribution. Instead, the overall shape of the distribution changes. The model we used to generate these distributions is structured as follows:

@@

where $\epsilon_{ijk}$ is normally distributed, and $\theta$ represents a threshold below which response values cannot occur. Unfortunately, ART's alignment mechanism is not designed to handle such cases. The alignment calculations shown in @fig-ART-explained include various mean terms that are sensitive to distribution differences across the levels of factors. In particular, the values of a random sample that appear in the tail of the right distribution can disproportionately influence the within-cell mean. This, in turn, leads to the exaggeration of all ranks within that cell, causing the method --- as our results demonstrate --- to confound effects.

We acknowledge that @elkin:2021 diverged from the modeling approach described by @higgins:1990 and @Mansouri:1995 and considered nonlinear models to evaluate ART. Yet, their experiments did not reveal that ART confounds effects, simply because they assessed Type I error rates only in scenarios where all main effects were null.

### Heteroscedasticity issues
@myers2012 explain that *"if the underlying distribution of the response variable is not normal, but is a continuous, skewed distribution, such as the lognormal, gamma, or Weibull distribution, we often find that the constant variance assumption is violated,* and in such cases, *"the variance is a function of the mean"* (Pages 54-55). This pattern frequently occurs in studies measuring task-completion times, whether the task is visual, motor, or cognitive. As tasks become more difficult and prolonged, variance tends to increase. Similarly, slower participants generally exhibit greater variance across tasks compared to faster, more practiced users. This suggests that ART's assumption of constant variance in skewed distributions is largely unrealistic.

@@

However, a question remains: Is ART’s failure in handling such models due solely to heteroscedasticity issues? For example, consider the two time distributions in @fig-homoscedastic-model, which --- despite having different shapes --- share the same standard deviation. Would ART perform correctly in this case? While we do not directly address this question in the main paper, we include additional results in the [appendix](appendix.html) showing that as long as distribution shapes differ across levels, ART tends to confound effects. Naturally, as the shapes become more similar, the problem diminishes. In conclusion, heteroscedasticity may exacerbate the problem, but it is not its sole cause.

@@

@fig-art-binary illustrates this problem using a simple dataset with a binary variable. In our example, all but one value are zero. Notice how dramatically ART alters the ranks and exaggerates the difference between the two levels of $B$ ($b_1$ vs. $b_2$). This issue persists even when a large number of participants are added and there is only a single value of 1 among thousands of 0s, causing the method to produce arbitrarily low *p*-values. Then, a single value variation (e.g., turning this 1 to 0) will cause ART’s results to drastically change. We will show that this unstable behavior makes ART inflate error rates even for variables with a larger range of discrete values.
